---
title: "Cyclistic Bike-Share: Strategy to Convert Casual Riders to Annual Members"
author: "Mr Matthew Ternenge Beeun"
date: "`r Sys.Date()`"
output: html_document
toc: true
toc_depth: 3
theme: spacelab
---

# load the packages into r studio
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(lubridate)
library(readxl)
library(rmarkdown)
```


# **Executive Summary**
This report summarizes the analysis of Cyclistic's historical trip data, focusing on behavioral differences between Annual Members and Casual Riders. The goal is to provide data-driven recommendations to maximize annual memberships, which are considered crucial for the company's future profitability.

The core finding is that Annual Members use the service as a utility for commuting, while Casual Riders use it for leisure and recreation.

## **Top 3 Actionable Recommendations (Act Phase):**

1. **Introduce a "Weekend Pass" or "Leisure Membership":** Target the casual riders' existing long, expensive weekend usage with a specialized, discounted membership to increase conversion with minimal friction.

2. **Implement Real-Time Cost-Saving Communication:** Use the mobile app to show casual riders their immediate, projected savings after a long, expensive trip to encourage immediate membership purchase.

3. **Geo-Target Promotions at Leisure Hubs:** Focus advertising efforts at high-casual-volume, tourist-centric stations during peak summer and weekend hours.


## import the data for cleaning and analysis
```{r Bike share}
divvy_2019_q1 <- read_excel("C:\\Users\\Mr Matthew\\Desktop\\portfolio\\Bike Share Case Study\\Datasets\\Divvy_Trips_2019_Q1.xlsx")

### second dataset
divvy_2020_q1 <- read_excel("C:\\Users\\Mr Matthew\\Desktop\\portfolio\\Bike Share Case Study\\Datasets\\Divvy_Trips_2020_Q1.xlsx")


```


## **1. Ask: Business Task & Key Questions**
The Director of Marketing, Lily Moreno, believes maximizing annual memberships is key to the company's success.

## **Business Task**

The objective is to analyze how annual members and casual riders use Cyclistic bikes differently, and use these insights to design a marketing strategy aimed at converting casual riders into annual members.

## Project Overview:
-- **Goal:** Understand how Casual Riders and Annual Members use Cyclistic Bikes                    differently.
-- **Stakeholders:** Director of Marketing: Lily Moreno, Cyclistic Marketing Analyst                Team, Cyclistic Executive Team.
-- **Data Sources:** Historical public data from Divvy, the name of the Chicago Bike                Share System operated by Lift. Data: divvy_2019_q1 and divvy_2019_q1
-- **Tools Used:** Adobe software for the file, Google Sheet, R-Studio

## **Guiding Questions**

1. How do annual members and casual riders use Cyclistic bikes differently?
2. Why would casual riders buy Cyclistic annual memberships?
3. How can Cyclistic use digital media to influence casual riders to become members?

## ** 2. Prepare & Process: Data Methodology**

The analysis used Cyclistic’s historical trip data for the past 12 months. This data is public and anonymized, ensuring privacy.

-- **Data Cleaning and Transformation**

The raw data required significant processing to ensure accuracy and consistency:

-- **Data Consolidation:**
Individual monthly files were merged into a single, comprehensive dataset.

-- **Data Consistency:**
Rider types were standardized: 'Subscriber' and 'member' were unified as 'Annual Member'; 'Customer' and 'casual' were unified as 'Casual Rider'.


-- **Feature Engineering:**

-- The **ride_length** (trip duration) was calculated from started_at and ended_at timestamps.

-- New fields for **day_of_week** and month were extracted to analyze temporal trends.

-- **Outlier Removal:** 
Rides with a duration of 0 seconds or rides exceeding 24 hours (likely quali**ty assurance or maintenance issues) were removed to prevent skewing the average duration metrics.


## **3. Analyze: Key Findings and Insights**
The descriptive analysis confirmed distinct differences in usage patterns, supporting the development of separate user personas.

## Cleaned and ready use data for analysis
```{r}
glimpse(davvy_clean_cyclistic_ready_data)
```

```{r}
# Example of creating a data frame to hold summary metrics

davvy_summary_stats_2 <- davvy_clean_cyclistic_ready_data %>%
  group_by(rider_type) %>%
  summarize(
    # 1. Total Number of Rides (Volume)
    number_of_rides = n(),
    
    # 2. Average Ride Length
    average_duration_mins = round(mean(ride_length), 4),
    
    # 3. Median Ride Length
    median_duration_mins = round(median(ride_length), 4),
    
    # 4. Maximum Ride Length
    max_duration_mins = round(max(ride_length), 4)
  )
print(davvy_summary_stats_2)
```

## Supporting visuals for the descriptive analysis key findings
```{r}
# Plotting a supporting visual for the key findings
ggplot(davvy_summary_stats_2, aes(x = rider_type, y = number_of_rides, fill = rider_type)) +
  # Create the bar chart
  geom_bar(stat = "identity", position = "dodge") + 
  
  # Add the value labels on top of the bars
  geom_text(aes(label = format(number_of_rides, big.mark = ",")), 
            vjust = -0.5, size = 4) +
  
  # Apply professional labels and title
  labs(
    title = "Total Number of Trips (Q1 2019 & 2020 Combined)",
    subtitle = "Annual Members use the service more frequently for routine travel.",
    x = "Rider Type",
    y = "Total Number of Rides",
    fill = "Rider Type"
  ) +
  # Clean up the theme for better readability
  theme_minimal() +
  # Rotate axis text if names are long
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5)) +
  # Remove the y-axis line to focus on the bar height
  scale_y_continuous(labels = scales::comma)


```

## Average usage comparisms
```{r}
ggplot(davvy_summary_stats_2, aes(x = rider_type, y = average_duration_mins, fill = rider_type)) +
  # Create the bar chart
  geom_bar(stat = "identity", position = "dodge") + 
  
  # Add the value labels on top of the bars, rounding to 1 decimal place for presentation
  geom_text(aes(label = round(average_duration_mins, 1)), 
            vjust = -0.5, size = 4) +
  
  # Apply professional labels and title
  labs(
    title = "Average Trip Duration by Rider Type",
    subtitle = "Casual Riders take trips that are significantly longer than Annual Members.",
    x = "Rider Type",
    y = "Average Ride Duration (Minutes)",
    fill = "Rider Type"
  ) +
  # Clean up the theme
  theme_minimal() +
  # Customize colors (optional, but good for reporting)
  scale_fill_manual(values = c("Casual Rider" = "orange", "Annual Member" = "steelblue"))


```
-- **Insight:** The disparity in average ride length strongly suggests members use the service for short commutes, while casual riders prioritize long-form leisure trips.

2. **Temporal Usage Patterns**
Analyzing the data by the day of the week reveals when each group is riding.

```{r}
# Assuming 'rider_type' and 'day_of_week' columns exist in 'clean_cyclistic_data'

davvy_daily_analysis <- davvy_clean_cyclistic_ready_data %>%
  group_by(rider_type, day_of_week) %>%
  summarize(
    total_rides = n(),
    average_duration_mins = round(mean(ride_length), 4)
  ) %>%
  ungroup() # Essential to remove grouping before the next step

print(davvy_daily_analysis)

```
## Supporting visuals for the temporal daily rides
```{r}
ggplot(davvy_daily_analysis, aes(x = day_of_week, y = total_rides, fill = rider_type)) +
  geom_col(position = "dodge") + # 'dodge' separates the bars for comparison
  labs(
    title = "Total Rides by Day of Week",
    subtitle = "Members use bikes for commuting (weekday focus); Casuals for leisure (weekend spike).",
    x = "Day of Week",
    y = "Total Rides",
    fill = "Rider Type"
  ) +
  theme_minimal() +
  scale_y_continuous(labels = scales::comma)

```



-- **Insight:** Members demonstrate a clear weekday commuting pattern, while casual riders' usage peaks significantly on the weekends.

### Average Ride Duration by Day of Week
```{r}
ggplot(davvy_daily_analysis, aes(x = day_of_week, y = average_duration_mins, fill = rider_type)) +
  geom_col(position = "dodge") + # 'dodge' separates the bars for comparison
  labs(
    title = "Average Duration by Day of Week",
    subtitle = "Casual Riders have Longer trips daily.",
    x = "Day of Week",
    y = "average_duration_mins",
    fill = "Rider Type"
  ) +
  theme_minimal() +
  scale_y_continuous(labels = scales::comma)

```

-- **Insight:** While casual riders take longer rides every day, their trip duration reaches its maximum on weekends, further solidifying the use case as primarily leisure and recreation.


## **4. Act: Top Three Recommendations**
The marketing strategy must target the unique behavioral characteristics of the Casual Rider (long, weekend, leisure use).

###**Recommendation 1: Introduce a "Weekend Pass" or "Leisure Membership"**
-- **Action:** Develop a new, attractively priced membership tier focused on recreational use, such as a "Cyclistic Weekend Warrior Pass" or a "Monthly Leisure Plan." This plan offers unlimited rides on Friday evening through Sunday for a low flat fee.
**Justification:** This meets the casual rider where they are. It converts their highest-volume, highest-cost activity (long weekend trips) into a cost-saving membership, providing an easy on-ramp to full annual membership.

###**Recommendation 2: Implement Real-Time Cost-Saving Communication**
-- **Action:** Use digital media (app notification or immediate post-ride email) to show casual riders the financial benefit of membership right after they complete an expensive, pay-per-ride trip.
**Justification:** By calculating the exact amount they would have saved on that specific trip with an annual membership (e.g., "You just spent $18. You could have saved $14 with a membership!"), we create an immediate, tangible incentive for conversion, leveraging the pain point of high occasional cost.

###**Recommendation 3: Geo-Targeted Promotions at Leisure Hubs**
--**Action:** Identify the top 5-10 start and end stations predominantly used by casual riders (often tourist attractions, parks, and entertainment districts) and dedicate enhanced marketing efforts at these locations.
**Justification:** Promotions should be deployed in the summer months and on weekends, utilizing eye-catching physical signage and location-aware digital ads that speak directly to the tourist/leisure mindset: "Explore longer, save more. Convert today!" This ensures the marketing message reaches the target audience at the optimal time and place.

## Conclusion
The analysis clearly demonstrates that Casual Riders operate on a recreational model, whereas Annual Members operate on a utility model. By strategically targeting the Casual Riders' preference for long, weekend-based leisure riding with tailored, value-driven membership products and conversion messaging, Cyclistic can effectively maximize the conversion rate and secure the more profitable Annual Member segment.